---
title: "Introduction: Stata to R"
author: "Lucy Hackett"
date: "7/8/2020"
output: 
  html_document:
    theme: yeti
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 4
    toc_float: yes
    keep_md: true
---

<style type="text/css">
body, td {
   font-size: 14px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Wecome to the introductory R Markdown document! In this short tutorial, we will take a shallow dive into the use of R (with a focus on the tidyverse) for reading, inspecting, cleaning and manipulating data. There are MANY excellent intros to R and data wrangling out there. To name a few:

* [Modern Dive](https://moderndive.com/index.html)
* [Tiago Ventura's Intro](https://tiagoventura.rbind.io/files/gsa_tidyverse_workshop)
* [D-Lab's workshops](https://dlab.berkeley.edu/training)

As well as hundreds of tutorials on DataCamp. I will not attempt to recreate these resources. Rather, view this guide as a "translation" of stata-ish syntax and logic to the R system. Therefore I will not go into detail about the different objects and syntax that I use here, but rather will work through practical examples of how things would be done in Stata vs. R

Also note that the flexibility of R allows for many potential ways to do things, making it difficult and probably ill-advised for me to create a one-to-one dictionary of Stata commands to R commands. While this can be overwhelming and confusing at first, with time I think you'll find that flexibility is actually a great feature that allows you to choose the right/most efficient/easiest method depending on your particular data structure or needs.

## Importing packages

In R we have what's called "base R" and there are packages. Base R has tons of capabilities for statistical and probability work, but packages expand this capability by providing us with more specialized commands for things like data visualization, data cleaning and Econometric models. In Stata, use of external packages is less prevalent, but you still may use external packages such as ```ivreg2``` or ```binscatter```, which you gain access to by a one-time installation via ```ssc install```. In R, we can do this either through the package manager window on the lower right side of RStudio, or through the ```install.packages()``` command. Once we have installed the package, in order to use it we call the following: 

```{r lib, results="hide", message=FALSE}
library(dplyr)
library(ggplot2)
library(haven)
```

## Directories, importing data 

When working in Stata, frequently the first line in any dofile is ``cd``. In R, we set our current directory with:

```{r cd, eval=FALSE}
setwd('/users/hackettl/drive_berkeley/GSI_GSR/2020_fall/tutorials')
```

```{r cd1, eval=TRUE, echo=FALSE}
setwd('/users/hackettl/drive_berkeley/GSI_GSR/2020_fall/tutorials/pandas_for_stata_users')
```

A great advantage of R over Stata is you can import many different kinds of files with realtive ease, and hold several datasets in memory at a time. Let's import a few files I have on hand.

**sinac_sample.dta** is a 1% random sample of birth registries from 2008 in Mexico. **cat_local_inegi.csv** is a dictionary of Mexican localities; it gives their name, state, municipality and population.

Here I use the most simple import, but like in insheet, you can choose to start/stop at certain csv rows, change the column names to lowercase, etc; for these options and many more, see the documentation.

Note that to read the birth data I use ``read_dta`` which is a command provided by the **haven** package.

```{r import1}
sinac <- read_dta('data/sinac_sample.dta')
local <- read.csv('data/cat_local_inegi.csv')
```

## Inspecting data

A good first step is always to take a look at the data; see how many (non-missing) observations we have, how the data behaves, etc. Here I show some examples of how to do this. In this section we will be mimicking Stata commands such as

* br
* sum
* bys: sum
* hist
* tab
* duplicates drop
* sort

I'll put the Stata command in a comment above the python command where possible, so if you're looking for a specific command, you can search this notebook for that command (i.e., search the document for "tab" for example.


```{r opt, echo=FALSE}
options(width = 100)
```

```{r br}
# br
head(sinac, row = 10)
```

I don't always love the information I can see from ``head()``, so if you want a more detailed look you can click on the dataset name in the Environment window (upper right in RStudio) to see a larger format of the data, though this isn't always bery efficient. You can also use this window to see a complete list of all the variables!

To get a quick look at some summary statistics of the data, you can use ``dfapply()`` which is an iterative function with the option "favstats":

```{r sum}
# sum sexo_nac
summary(sinac$sexo_nac)
```

Note that in R, we have to associate variables we are interested in with the dataset they belong to, because we may and in fact do in this case have several datasets in memory. To do this, we use the syntax ``dataset_name$variable_name``. In the tidyverse and in other packages, we can use the pipe ``%>%`` to pass identifying information or transformations onto the next operation:

```{r tab}
# tab atendio
sinac %>% count(atendio)
```


```{r drop_dup}
# duplicates drop, save to different data
sinac_unique <- sinac %>% distinct()
```


```{r bys}
# to do operations by groups, like bys: sum 
# (you can add more variables to the list at will)
# note that we need to ignore missing values with na.rm = TRUE
sinac %>% group_by(ent_res_cve) %>% 
  summarise(avg_weight = mean(peso_nac_vivo, na.rm = TRUE),
            min_weight = min(peso_nac_vivo, na.rm = TRUE))

```

Be careful saving data as grouped, as this may affect future modifications you make to the data in unexpected ways!

```{r sort}
# view the data sorted: sort
sinac %>% arrange(edad_madre,apgar)
# again, here I haven't saved this, I'm just looking at it. 
# To save this, set it equal to a name

```

Looks like the youngest mothers in our data are 9 years old. Let's plot mother's age to get a sense of the distribution:

```{r hist}

hist(sinac$edad_madre)

```

Here's a nicer version using ``ggplot``:

```{r gghist, message=FALSE}

ggplot(data = sinac, aes(x = edad_madre)) +
  geom_histogram(color="darkblue", fill="lightblue") + 
  xlab("Mother's age") + 
  theme_bw()
  
```

##  A note about object assignment
A theme you may have noted above is that we can do things to the data without changing the underlying data or even saving what we're doing. This may remind you of what you might do in Stata with ``preserve ... restore``, which is the only way in Stata (besides tempfiles I guess) to explore manipulations of the data without altering the underlying data.

This is one of the main advantages that R has over Stata; we can hold multiple data, or multiples versions/ manipulations of the data in memory, leaving our original dataset intact! 

## Data cleaning 

Now we've looked around a bit at the data, we're going to clean it. These type of commands in Stata are:

* rename
* replace
* gen
* collapse
* merge

First let's get these column names in English...

```{r rename_all}
# rename
# note that here I AM reassigning the data to save changes
sinac <- sinac %>% rename(mother_age = edad_madre,
                           no_preg = numero_embarazos,
                           length = talla_nac_vivo,
                           weight = peso_nac_vivo,
                           no_appts = tot_consult,
                           day_born = dia_nac_hijo,
                           mo_born = mes_nac_hijo,
                           yr_born = year_nac_hijo, 
                           sex = sexo_nac, 
                           gest_age = semanas_gest, 
                           type_doc = atendio)

sinac %>% head()
```

```{r replace}
# Let's change the state code from 99 to missing
# replace
# ifelse has arguments condition, what to do if true, then what to do if false
sinac <- sinac %>% mutate(ent_res_cve = 
                            ifelse(
                              ent_res_cve == 99,
                              NaN,
                              ent_res_cve
))

summary(sinac$ent_res_cve)
# good! looks like the max code is now 32 (the number of states in Mexico)
```


```{r gen}
# gen
# Now let's make a categorical variable for mother's schooling 
sinac <- sinac %>% mutate(schooling_cat = 
                           ifelse(
                             schooling_mother < 4,
                             1,
                             ifelse(
                               schooling_mother >= 4 & schooling_mother <= 6,
                               2,
                               3)
                           ))

summary(sinac$schooling_cat)

```
